Discrete Quantum Theories and Computing

### 2018-03-19 dissertation.tex
dissertation.lyx "3.4 Discrete Quantum Theory (I)" Updating...

\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{}[section]

@@

Our next objective is to develop more realistic discrete quantum theory
variants that exclude ``supernatural'' algorithms such as the one
presented above. Our first such plausible framework~\cite{HOSW2011}
is based on complexifiable finite fields. To incorporate complex numbers
for quantum amplitudes, we exploit the fact that the polynomial $x^{2}+1=0$
is \emph{irreducible} (has no solution) over a prime field $\ff{p}$
with $p$ odd if and only if $p$ is of the form $4\ell+3$, with
$\ell$ a non-negative integer. In other words, $x^{2}+1=0$ is irreducible
over $\ff{3}$, $\ff{7}$, $\ff{11}$, $\ff{19}$, \ldots{}. We achieve
our goal by observing that any field in this family is extensible
to a field~$\ff{p^{2}}$ whose elements can be viewed as discrete
complex numbers with the real and imaginary parts in~$\ff{p}$. In
a complexified finite field $\Fpp$, the Frobenius automorphism that
maps $\alpha\in\Fpp$ to $\alpha^{p}\in\Fpp$ acts like complex conjugation
\cite{Artin1991,grove2002classical,DummitFoote2004}. For example,
in $\Fppx{3}$, we have $\left(2+\rmi\right)^{3}=8+12\rmi-6-\rmi=2+11\rmi$
which, in the field, is equal to $2-\rmi$ since $11\equiv-1\pmod{3}$.

We define the \emph{field norm} $\fnorm{\cdot}$ as the map from $(a+ib)\in\Fpp$
to $a^{2}+b^{2}\in\Fp$, 
\begin{equation}
\fnorm{\alpha=a+\rmi b}=a^{2}+b^{2}\,.\label{fieldnorm.eq}
\end{equation}
We avoid the square root in the discrete field framework because,
unlike the continuous case, the square root does not always exist.

@@

In this section we want to build a theory of discrete vector spaces
that approximates as closely as possible the features of conventional
quantum theory. Such a structure would ideally consist of the following:
(i) a vector space over the field of complex numbers, and (ii) an
inner product $\ip{\Phi}{\Psi}$ associating to each pair of vectors
a complex number, and satisfying the following properties: 
\begin{itemize}
\item[A.] $\ip{\Phi}{\Psi}$ is the complex conjugate of $\ip{\Psi}{\Phi}$; 
\item[B.] $\ip{\Phi}{\Psi}$ is conjugate linear in its first argument and
linear in its second argument; 
\item[C.] $\ip{\Psi}{\Psi}$ is always non-negative and is equal to~0 only
if $\ket{\Psi}$ is the zero vector. 
\end{itemize}
It turns out that a vector space defined over a finite field cannot
have an inner product satisfying the properties above. However, we
will introduce an Hermitian ``dot product'' satisfying some of those
properties.

We are interested in the $n$-qubit vector space ${\cal H}$ of dimension
$D=2^{n}$ defined over the complexified field $\Fpp$. Let $\ket{\Psi}=(\alpha_{0}~\alpha_{1}~\ldots~\alpha_{D-1})^{T}$
and $\ket{\Phi}=(\beta_{0}~\beta_{1}~\ldots~\beta_{D-1})^{T}$ represent
vectors in ${\cal H}$, with numbers $\alpha_{i}$ and $\beta_{i}$
drawn from $\Fpp$, and where $(\cdot)^{T}$ is the transpose.

\begin{definition}[Hermitian dot product] Given vectors $\ket{\Phi}$
and $\ket{\Psi}\in{\cal H}$, the Hermitian dot product of these vectors
is: 
\begin{equation}
\ip{\Phi}{\Psi}=\sum_{i=0}^{D-1}~\beta_{i}^{p}~\alpha_{i}^{\;}\ .
\end{equation}
\end{definition} Two vectors $\ket{\Phi}$ and $\ket{\Psi}\in{\cal H}$
are said to be orthogonal if $\ip{\Phi}{\Psi}=0$. This product satisfies
conditions A and B for inner products but violates condition C since
in every finite field there always exists a non-zero vector $\ket{\Psi}$
such that $\ip{\Psi}{\Psi}=0$. The reason is that addition in finite
fields eventually ``wraps around'' (because of their cyclic or modular
structure), allowing the sum of non-zero elements to be zero. The
fraction of non-zero vectors satisfying $\ip{\Psi}{\Psi}=0$ decreases
with the order $p$.

For any vector $\ket{\Psi}=(\alpha_{0}~\alpha_{1}~\ldots~\alpha_{D-1})^{T}$,
the Hermitian dot product $\ip{\Psi}{\Psi}$ is equal to $\sum_{i=0}^{D-1}~\fnorm{\alpha_{i}}$,
which is the sum of the field norms for the complex coefficients.
For convenience, we now extend the field norm to include vector arguments
by defining 
\begin{equation}
\fnorm{\ket{\Psi}}=\ip{\Psi}{\Psi}=\sum_{i=0}^{D-1}~\fnorm{\alpha_{i}}\ .\label{fnormD.eq}
\end{equation}
The field norm of a vector can vanish for non-vanishing vectors.

>>>>

\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

@@

Our next objective is to develop more realistic discrete quantum theory
variants that exclude ``supernatural'' algorithms such as the one
presented above. Our first such plausible framework~\cite{HOSW2011}
is based on complexifiable finite fields. To incorporate complex numbers
for quantum amplitudes, we exploit the fact that the polynomial $x^{2}+1$
is \emph{irreducible} ($x^{2}+1=0$ has no solution) over a prime
field $\ff{p}$ with $p$ odd if and only if $p$ is of the form $4\ell+3$,
with $\ell$ a non-negative integer \cite{Artin1991,DummitFoote2004}.
For example, given any $x\in\ff{3}=\left\{ 0,\pm1\right\} $, when
$x=0$, we have $x^{2}+1=0^{2}+1=1$; when $x=\pm1$, we have $x^{2}+1=\left(\pm1\right)^{2}+1=1$.
Therefore, $x^{2}+1$ is irreducible over $\ff{3}$. In contrast,
over $\ff{5}$, $2^{2}+1=0$ so that $x^{2}+1$ is reducible.

We achieve our goal by observing that any field $\ff{p}$ with $p=4\ell+3$
is extensible to a field~$\ff{p^{2}}$ whose elements can be viewed
as discrete complex numbers with the real and imaginary parts in~$\ff{p}$.
In a complexified finite field $\Fpp$, the Frobenius automorphism
that maps $\alpha\in\Fpp$ to $\alpha^{p}\in\Fpp$ acts like complex
conjugation \cite{Artin1991,grove2002classical,DummitFoote2004}.
For example, in $\Fppx{3}$, we have $\left(2+\rmi\right)^{3}=8+12\rmi-6-\rmi=2+11\rmi$
which, in the field, is equal to $2-\rmi$ since $11\equiv-1\pmod{3}$.

We define the \emph{field norm} $\fnorm{\cdot}$ as the map from $a+\rmi b\in\Fpp$
to $a^{2}+b^{2}\in\Fp$, 
\begin{equation}
\fnorm{a+\rmi b}=a^{2}+b^{2}\,.\label{fieldnorm.eq}
\end{equation}
We avoid the square root in the discrete field framework because,
unlike the continuous case, the square root does not always exist.

@@

In this section we want to build a theory of discrete vector spaces
that approximates as closely as possible the features of conventional
quantum theory. Such a structure would ideally consist of the following:
(i) a vector space over the field of complex numbers, and (ii) an
inner product $\ip{\Phi}{\Psi}$ associating to each pair of vectors
a complex number, and satisfying the following properties: 
\begin{enumerate}[label=(\Alph{enumi})]
\item $\ip{\Phi}{\Psi}$ is the complex conjugate of $\ip{\Psi}{\Phi}$; 
\item $\ip{\Phi}{\Psi}$ is conjugate linear in its first argument and linear
in its second argument; 
\item $\ip{\Psi}{\Psi}$ is always non-negative and is equal to~0 only
if $\ket{\Psi}$ is the zero vector. 
\end{enumerate}
It turns out that a vector space defined over a finite field cannot
have an inner product satisfying the properties above. However, we
will introduce an Hermitian ``dot product'' satisfying some of those
properties.

We are interested in the vector space ${\cal H}$ of dimension $D$
defined over the complexified field $\Fpp$. Let $\ket{\Psi}=\begin{pmatrix}\alpha_{0} & \alpha_{1} & \ldots & \alpha_{D-1}\end{pmatrix}^{T}$
and $\ket{\Phi}=\begin{pmatrix}\beta_{0} & \beta_{1} & \ldots & \beta_{D-1}\end{pmatrix}^{T}$
represent vectors in ${\cal H}$, with numbers $\alpha_{i}$ and $\beta_{i}$
drawn from $\Fpp$, and where $\left(\cdot\right)^{T}$ is the transpose.

\begin{definition}[Hermitian dot product] Given vectors $\ket{\Phi}$
and $\ket{\Psi}\in{\cal H}$, the Hermitian dot product of these vectors
is: 
\begin{equation}
\ip{\Phi}{\Psi}=\sum_{i=0}^{D-1}\beta_{i}^{p}\alpha_{i}\,.
\end{equation}
\end{definition} Two vectors $\ket{\Phi}$ and $\ket{\Psi}\in{\cal H}$
are said to be orthogonal if $\ip{\Phi}{\Psi}=0$. This product satisfies
conditions A and B for inner products but violates condition C since
in every finite field there always exists a non-zero vector $\ket{\Psi}$
such that $\ip{\Psi}{\Psi}=0$. The reason is that addition in finite
fields eventually ``wraps around'' (because of their cyclic or modular
structure), allowing the sum of non-zero elements to be zero. The
fraction of non-zero vectors satisfying $\ip{\Psi}{\Psi}=0$ decreases
with the order $p$.

For any vector $\ket{\Psi}=\begin{pmatrix}\alpha_{0} & \alpha_{1} & \ldots & \alpha_{D-1}\end{pmatrix}^{T}$,
the Hermitian dot product $\ip{\Psi}{\Psi}$ is equal to $\sum_{i=0}^{D-1}\fnorm{\alpha_{i}}$,
which is the sum of the field norms for the complex coefficients.
For convenience, we now extend the field norm to include vector arguments
by defining 
\begin{equation}
\fnorm{\ket{\Psi}}=\ip{\Psi}{\Psi}=\sum_{i=0}^{D-1}\fnorm{\alpha_{i}}\,.\label{fnormD.eq}
\end{equation}
The field norm of a vector can vanish for non-vanishing vectors.
